Designing Types for R, Empirically

### 2020-10-15 paper/main.tex
minor

Given the pecularities of the language there are a number of design
choices that need to be reviewed and evaluated. It is uncontroversial
to include types that cover the six kinds of primitive vectors,
furthermore environments and the distinguished null types are commonly
used and must be included. For simplicity, we omit some of the data
types that closer to the implementation of the
language. Fig.~\ref{types} presents our type language.

\begin{figure}[!h] \noindent \small  \centering\begin{minipage}{.45\linewidth}
\begin{tabular}{lclr}
\T& ::= & \any          & \it top type\\
  & |   & \nul         & \it null type\\
  & |   & \environment  & \it environment type \\
  & |   & \S            & \it scalar type \\
  & |   & \V            & \it vector type \\
  & |   & \T~\k{|}\T    & \it union type \\
  & |   & \k{?} \T      & \it nullable type \\
  & |   & \FUN{A_1, \dots~ A_n}\T    & \it function type \\
  & |   & \LIST\T                  & \it list type\\
  & |   & \CLASS{ID_1, \dots~ ID_n}  & \it class type\vspace{5pt}\\
\end{tabular}\end{minipage}   \hfill
\begin{minipage}{.45\linewidth}\begin{tabular}{lclr}
\A & ::= & \T          & \it arguments \\
   & |   & \VARGS      & \it dots \vspace{5pt} \\
\V & ::= & ~\VEC\S     & \it vector types\\
   & |   & \NAVEC\S    & \it na vector types \vspace{5pt}\\
\S & ::= & \int        & \it integer\\
   & |   & \chr        & \it character\\
   & |   & \dbl        & \it double \\
   & |   & \lgl        & \it logical\\
   & |   & \clx        & \it complex \\
   & |   & \raw        & {\it raw } \vspace{5pt}\\%
\end{tabular}\end{minipage}\caption{The R type language}\label{types}
\end{figure}

\paragraph{Scalars} While R does not have scalar data types, there are
cases where functions except scalar values, for example a conditional
takes a single logical (and will complain if more values are passed).
We considered tracking dimensions of data structures, but decided
against it. Instead, the type language differentiates between vectors
of lenght 1 and vectors of any dimension.  The primitive types can be
either vectors (e.g., \VEC\int) or scalars (e.g., \int).  A vector can
happen to be of length 1, and thus a scalar is also a vector.  Vectors
are monomorphic, a vector of doubles contains only doubles.

\paragraph{Missing} 
Each of basic types has its specific \NA. Many built-in functions,
especially those implemented in C or Fortran, do not support \NA
values.  It is thus advantageous to distinguish between vector that
can contain missing values and those that are guaranteed not to.  In
our experience, functions that expect scalar values tend to not admit
{\NA}s, thus scalar types are treated as being \NA-free. The type
language does allows to write \VEC\int to specify that a vector of
integers may have {\NA}s and \NAVEC\int to say that a vector must not
have missing values. Of course, a value of type \VEC\int may happen to
not have any \NA and thus be of type \NAVEC\int.  The type \raw does
not allow {\NA}s, so \VEC\raw is \NA-free.

@@

\begin{table}[H]
	\centering
	\resizebox{.98\linewidth}{!}{
		\begin{tabular}{llrrr}
			\toprule
			Merge Strategy & Function Type \\
			\midrule
			Union of Arrows &  \makecell[tl]{
		      \FUN{\CLASS{{\tt data.frame}}, ...}{\CLASS{{\tt data.frame}}}	| \\
             \FUN{\CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}}, ...}{\CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}}} | \\
             \FUN{\CLASS{{\tt data.frame, tbl, tbl\_df}}, ...}{\CLASS{{\tt data.frame, tbl, tbl\_df}}}
			} \\
		\addlinespace
			Hybrid & \makecell[tl]{
				\FUN{\CLASS{{\tt data.frame}}, ...}{\CLASS{{\tt data.frame}}}	| \\
				\FUN{\CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}}, ...}{\CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}}} | \\
				\FUN{\CLASS{{\tt data.frame, tbl, tbl\_df}}, ...}{\CLASS{{\tt data.frame, tbl, tbl\_df}}}
			} \\
		\addlinespace
			Arrow of Unions & \makecell[tl]{
				$\langle \CLASS{{\tt data.frame}} \;|\; \CLASS{{\tt data.frame, tbl, tbl\_df}} \;|\; \CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}, ...} \rangle$ $\rightarrow$ \\
				$\CLASS{{\tt data.frame}} \;|\; \CLASS{{\tt data.frame, tbl, tbl\_df}} \;|\; \CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}}$
			} \\
			\bottomrule
	\end{tabular}}
	\caption{Types of \code{X::Y} with different type merge strategies. }\label{tbl:real-function-different-merge-strats-2}
\end{table}

>>>>

Given the pecularities of the language there are a number of design
choices that need to be reviewed and evaluated. It is uncontroversial
to include types that cover the six kinds of primitive vectors,
furthermore environments and the distinguished null types are commonly
used and must be included. Fig.~\ref{types} presents our type language.

\begin{figure}[!h] \noindent \small  \centering\begin{minipage}{.45\linewidth}
\begin{tabular}{lclr}
\T& ::= & \any          & \it top type\\
  & |   & \nul         & \it null type\\
  & |   & \environment  & \it environment type \\
  & |   & \S            & \it scalar type \\
  & |   & \V            & \it vector type \\
  & |   & \T~\k{|}\T    & \it union type \\
  & |   & \k{?} \T      & \it nullable type \\
  & |   & \FUN{A_1, ..., A_n}\T    & \it function type \\
  & |   & \LIST\T                  & \it list type\\
  & |   & \CLASS{ID_1, ..., ID_n}  & \it class type\vspace{5pt}\\
\end{tabular}\end{minipage}   \hfill
\begin{minipage}{.45\linewidth}\begin{tabular}{lclr}
\A & ::= & \T          & \it arguments \\
   & |   & \VARGS      & \vspace{5pt} \\
\V & ::= & ~\VEC\S     & \\
   & |   & \NAVEC\S    & \it na vector types \vspace{5pt}\\
\S & ::= & \int        & \\
   & |   & \chr \\
   & |   & \dbl \\
   & |   & \lgl \\
   & |   & \clx \\
   & |   & \raw \vspace{5pt}\\
\end{tabular}\end{minipage}\caption{The R type language}\label{types}
\end{figure}

\paragraph{Scalars} We distinguish between
vectors of length 1, and vectors of any dimension. The base types, \int,
\dbl, \clx, \lgl (booleans), \chr, and \raw (bytes), can be either vectors
(e.g., \VEC\int) or scalars (e.g., \int).  A vector can happen to be of
length 1, and thus a scalar value is also a vector.  Vectors are
homogeneous, in that a vector of doubles contains only doubles.

\paragraph{Missing} 
Each of basic types has its specific \NA. Thus there is an \int \NA as
well as a \dbl \NA, and they are different.  The type system allows
one to distinguish between vectors that may contain {\NA}s (written
\NAVEC\int) and those who are guarantee to be \NA-free (\VEC\int).  A
\NA-free vector can be treated as a vector that may {\NA}s.  We do not
allow scalar values to be \NA and the type \raw does not allow {\NA}s.

@@

\begin{table}[H]
	\centering
	\resizebox{.98\linewidth}{!}{
		\begin{tabular}{llrrr}
			\toprule
			Merge Strategy & Function Type \\
			\midrule
			Union of Arrows &  \makecell[tl]{
		      \FUN{\CLASS{{\tt data.frame}}, ...}{\CLASS{{\tt data.frame}}}	| \\
             \FUN{\CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}}, ...}{\CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}}} | \\
             \FUN{\CLASS{{\tt data.frame, tbl, tbl\_df}}, ...}{\CLASS{{\tt data.frame, tbl, tbl\_df}}}
			} \\
		\addlinespace
			Hybrid & \makecell[tl]{
				\FUN{\CLASS{{\tt data.frame}}, ...}{\CLASS{{\tt data.frame}}}	| \\
				\FUN{\CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}}, ...}{\CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}}} | \\
				\FUN{\CLASS{{\tt data.frame, tbl, tbl\_df}}, ...}{\CLASS{{\tt data.frame, tbl, tbl\_df}}}
			} \\
		\addlinespace
			Arrow of Unions & \makecell[tl]{
				$\langle \CLASS{{\tt data.frame}} \;|\; \CLASS{{\tt data.frame, tbl, tbl\_df}} \;|\; \CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}, ...} \rangle$ $\rightarrow$ \\
				$\CLASS{{\tt data.frame}} \;|\; \CLASS{{\tt data.frame, tbl, tbl\_df}} \;|\; \CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}}$
			} \\
			\bottomrule
	\end{tabular}}
	\caption{Types of \code{dplyr::rename} with different type merge strategies. }\label{tbl:real-function-different-merge-strats-2}
\end{table}
